---
title: "Open Redcedar Analyses Instructions"
author: "Contributors: Joey Hulbert,..."
output: html_document
---

|            |            |            |            |
|:----------:|:----------:|:----------:|:----------:|
|[Redcedar](https://jmhulbert.github.io/open/redcedar)|[Data](https://jmhulbert.github.io/open/redcedar/data)|[ Analyses](https://jmhulbert.github.io/open/redcedar/analyses)|[Instructions](https://jmhulbert.github.io/open/redcedar/instructions)|
|             |           |            |            |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

We will use a Github Repository to host Github Pages written with R Markdown to document methods in analyses. 

### Set Up

* Getting set up
  + Install R and Rstudio
  + Create a login on github
  + Install git. Optionally install GitHub Desktop. Rstudio also has built in github functionality, but you may need to install git on your computer. 
  + Clone the [Github Repository](https://github.com/jmhulbert/open) to your computer.
  + Open Rstudio project 'open.Rproj' in rStudio.

### Github and R Markdown

We will document methods with R Markdown. Please create a R Markdown document to track methods even if the analyses are not completed in R (e.g. with ArcGIS).
  
* Creating a github page to document methods
  + Browse to analyses (open/redcedar/analyses) folder
  + Create a new folder for your analysis (e.g. cluster)
  + Create a html markdown document in the new folder
  + List the contributors to the analysis in the Rmarkdown YAML
  + Copy top table from other index.Rmd files to include navigation pane (redcedar, data, analyses, etc))
  + Clear default text if desired
  + Add a 'welcome or purpose header and description'
  + Save the markdown document as index.Rmd
  + Knit the document into index.html
  + Commit and Push changes to github repository
    + Describe changes or progress for other collaborators as commit message.

### Completing analyses
  
* Completing analyses 
  + Please complete analyses using data in the [data folder](https://jmhulbert.github.io/open/redcedar/analyses)
    + avoid copying the data into new locations, lets try to work from the same dataset throughout.
    + data can be downloaded [here](https://github.com/jmhulbert/open/raw/main/redcedar/data/data-modified.csv) or read inline using the command below.
  + Please document all steps throughout methods so they can be easily reproduced by novice data scientist.
  + Use the 'reclassified.tree.canopy.symptoms' column as the response variable when possible, however, use the 'binary.tree.canopy.symptoms' column as the response variable when there is not enough resolution to split data into 5 groups. 

```{r eval=FALSE, include=TRUE}
data <- read.csv(https://github.com/jmhulbert/open/raw/main/redcedar/data/data-modified.csv)
```



* File storage
  + Please save R analyses and related (small) datasets in the github repository.
  + All other files (e.g. GIS files) can be stored locally and shared later. 
  
### Sharing analyses and methods
  
Link to your analysis page from the main [active analyses page](https://jmhulbert.github.io/open/redcedar/analyses) when ready.   
  
* Create link in analyses page
  + Browse analyses folder and open index.Rmd file
  + Add the link to your new analysis folder/page
  + Add your name as a contributor in the Rmarkdown YAML
  + Knit the document into index.html
  + Commit and push changes to github repository.
    + Describe changes or progress for other collaborators as commit message.


