---
title: "gssurgo data analysis"
author: 'Contributors: Brendan Farrell, Joey Hulbert,...'
date: "2023-06-09"
output: html_document
---

|            |            |            |            |
|:----------:|:----------:|:----------:|:----------:|
|[Redcedar](https://jmhulbert.github.io/open/redcedar)|[Data](https://jmhulbert.github.io/open/redcedar/data)|[ Analyses](https://jmhulbert.github.io/open/redcedar/analyses)|[Instructions](https://jmhulbert.github.io/open/redcedar/instructions)|
|             |           |            |            |


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE,warning=FALSE}
library(tidyverse)
#devtools::install_github("lhmrosso/XPolaris")
library(randomForest)
library(caret)
library(rpart)
library(knitr)
library(corrplot)
```


# Approach

Brendan Farrell of Clockwork Micro extracted gssurgo soils data for the [data](https://jmhulbert.github.io/open/redcedar/data) using mixed methods as described below. 

* Three tables from gssurgo data were joined with iNat data
  + muaggat
  + component 


Then we completed a random forest analysis with the gssurgo soils data to see what variables were the most important for classifying trees as healthy or unhealthy. 


# Data Wrangling

### Import iNat Data - Empirical Tree Points (Response variables)

The steps for wrangling the [data](https://jmhulbert.github.io/open/redcedar/data) are described [here](https://jmhulbert.github.io/open/redcedar/data).**More details coming soon.** 

In general, Brendan used a combination of qGIS and PostGIS to extract soils data for the 2223 trees in the [data](https://jmhulbert.github.io/open/redcedar/data). 


Brendan joined datasets so iNat fields were already present in gssurgo dataset provided. 

```{r}
#data <- read.csv('https://github.com/jmhulbert/open/raw/main/redcedar/data/data-modified.csv')
#data <- read.csv('~/ServerFiles/open/redcedar/data/data-modified.csv')
#data$id <- as.character(data$id)
```


```{r}
gssurgo <- read.csv('~/ServerFiles/open/redcedar/data/datamatched20230620-resaved.csv')
```

There were a few rows that were not seperating correctly because of symbols in the description or field notes. We need to come abck to them, but for now we dropped the following observations

* ID 
  + "85409524" 
  + "108776647" 
  + "127848282"

```{r}
gssurgo <- gssurgo %>% filter(id!="85409524" & id!="108776647" & id!="127848282")
```


```{r}
gssurgo$binary.tree.canopy.symptoms <- as.factor(gssurgo$binary.tree.canopy.symptoms)
```


# Random Forests Analyses

More information coming soon

# Data Visualization

```{r}
ggplot(gssurgo,aes(wtdepaprjunmin,fill=binary.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```


