---
title: "gssurgo data analysis"
author: 'Contributors: Joey Hulbert,...'
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
---

|            |            |            |            |
|:----------:|:----------:|:----------:|:----------:|
|[Redcedar](https://jmhulbert.github.io/open/redcedar)|[Data](https://jmhulbert.github.io/open/redcedar/data)|[ Analyses](https://jmhulbert.github.io/open/redcedar/analyses)|[Instructions](https://jmhulbert.github.io/open/redcedar/instructions)|
|             |           |            |            |


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE,warning=FALSE}
library(tidyverse)
library(randomForest)
library(caret)
library(rpart)
library(knitr)
library(corrplot)
library(soilDB)
library(ggmap)
```


# Approach

A [previous analysis](./archived/) with ssurgo data collected through a combination of qGIS and PostGIS was completed and archived. 

The approach of the analysis on this page was to explore methods for integrating ssurgo data with r commands. 

Unfortunately, working with the soilDB package requires queries to interact with the 'soil data access' SDA web service using a language similar to SQL. Therefore, the below code may not look like r code. 

> TIMEOUT - pausing this analysis :)

* Resources
  + [Soil Data Access tutorial](https://ncss-tech.github.io/AQP/soilDB/SDA-tutorial.html)
  + [Tutorials to extract data with R program](https://sdmdataaccess.nrcs.usda.gov/QueryHelp.aspx)
  + [Soil data access query help](https://nasis.sc.egov.usda.gov/NasisReportsWebSite/limsreport.aspx?report_name=SDA-SQL_Library_Home)

# Data Wrangling

## Import iNat Data - Empirical Tree Points (Response variables)

The steps for wrangling the [data](https://jmhulbert.github.io/open/redcedar/data) are described [here](https://jmhulbert.github.io/open/redcedar/data).


```{r}
#data <- read.csv('https://github.com/jmhulbert/open/raw/main/redcedar/data/data-modified.csv')
data <- read.csv('~/ServerFiles/open/redcedar/data/data-modified.csv') #ofline
#data$id <- as.character(data$id)
```

## Prepare iNat data for soilDB

```{r}
#names(data[c(1:30)])
#summary(as.factor(data$place_country_name))
```

Filter data to observations in USA

```{r}
data <- data %>% filter(place_country_name=="United States")
```


```{r eval=FALSE, include=FALSE}
#map <- get_map(bbox = c(left = min(data$longitude), bottom = min(data$latitude), right = max(data$longitude),                  top = max(data$latitude)), zoom = 2)
```

```{r}
ggplot(data) +
  geom_point(aes(x = longitude, y = latitude), color = "blue") +
  xlab("Longitude") +
  ylab("Latitude") +
  theme_minimal()
```


```{r}
gps <- data[c(2,24,25)] #subset data to only include id and gps coordinates
gps <- gps %>% `colnames<-`(c("id","lat","lon"))
#gps$lat <- round(gps$lat,5)
#gps$lon <- round(gps$lon,6)
```

```{r}
# test dataset
coordinates <- data.frame(
  lat = c(35.2812, 35.2965, 35.4237),
  lon = c(-120.6766, -120.8565, -120.7661)
)
```


```{r}
# need to figure out how to use fetchSDA correctly
# gssurgo <- fetchSDA_spatial(coordinates)
?fetchSDA_spatial
#gssurgo <- get_component_from_SDA(gps)
```

