---
title: "Random Forests - Topographic Data Analysis"
author: 'Contributors: Joey Hulbert,...'
output: html_document
---

|            |            |            |            |
|:----------:|:----------:|:----------:|:----------:|
|[Redcedar](https://jmhulbert.github.io/open/redcedar)|[Data](https://jmhulbert.github.io/open/redcedar/data)|[ Analyses](https://jmhulbert.github.io/open/redcedar/analyses)|[Instructions](https://jmhulbert.github.io/open/redcedar/instructions)|
|             |           |            |            |


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE,warning=FALSE}
library(tidyverse)

library(randomForest)
library(caret)
library(rpart)
library(knitr)
library(corrplot)
```

```{r}
library(sf)
library(elevatr)
library(spatialEco)
library(geodata)
```


# Approach

The overall approach is to model empirical data collected by community scientists with ancillary topo data to identify important predictors of western redcedar dieback.

## Data Wrangling

### Import iNat Data - Empirical Tree Points (Response variables)

The steps for wrangling the iNat [data](https://jmhulbert.github.io/open/redcedar/data) are described [here](https://jmhulbert.github.io/open/redcedar/data).

```{r}
#data <- read.csv('https://github.com/jmhulbert/open/raw/main/redcedar/data/data-modified.csv')

data <- read.csv("~/ServerFiles/open/redcedar/data/data-modified.csv") #offline
```

### Spatial Data

```{r}
wgs84 <- "EPSG:4326"
utm10 <- "EPSG:26910"
```

```{r}
setwd("~/ServerFiles/open/redcedar/analyses/random-forests/topo")
dat.dir <- file.path(getwd(),"spatial.data")
data.dir.ex <-file.path("~/Volumes/TimeSync2TB/Data/Postdoc/R/Redcedar/Topo")
```

#### Country Boundary



```{r}
## hashtagged out because files already created previously
#bdy <- st_as_sf(gadm("USA",level=0,path=tempdir(),version="latest",resolution=1))
#geo.ext <- ext(bdy)
#bdy <- st_transform(bdy,utm10)
#st_write(bdy,file.path(dat.dir,"bdy.shp"))
```



```{r}
bdy <- st_read(file.path(dat.dir,"bdy.shp"))
#geo.ext.utm <- ext(426491.16359116934, 4539238.107476624, 720515.3028644782, 5316783.940854039,xy=TRUE) #project extent in EPSG:26910 decimals set to only include WA and OR.
```

#### 100m reference raster in UTM10 projection

```{r}
ref <- rast(ext(bdy),resolution=c(100,100),crs=crs(utm10))
```

#### x,y coordinate rasters

```{r}
#xcoord <- ref
#xcoord[] <- xyFromCell(ref, 1:ncell(ref))[,1]
#xcoord <- mask(xcoord, vect(bdy)) 
#names(xcoord) <- "xcoord"
#writeRaster(xcoord, file.path(dat.dir, "xcoord.tif")) 

#ycoord <- ref
#ycoord[] <- xyFromCell(ref, 1:ncell(ref))[,2]
#ycoord <- mask(ycoord, vect(bdy)) 
#names(ycoord) <- "ycoord"
#writeRaster(ycoord, file.path(dat.dir, "ycoord.tif")) 
```

#### Elevation (100m or 30 arc sec)

```{r}
elev <- elevation_30s(country="USA", path=tempdir())
geo.ext <- ext(-125.0000, 43.000, -113.0000, 50.0000,xy=TRUE) # project bounds data in EPSG:4326 decimals set to only include WA and OR. Also included xy=TRUE because they're in (ymin,xmin,ymax,xmax) order. 
elev <- crop(elev,geo.ext)
elev <- mask(project(elev, ref, method="bilinear"), vect(bdy))
names(elev) <- "elev"
writeRaster(elev, file.path(dat.dir.ex, "elev.tif"),overwrite=TRUE)  
```


#### Heat Load Index

```{r}
heat.load <- hli(elev)
names(heat.load) <- "hli"
writeRaster(heat.load, file.path(dat.dir.ex, "hli.tif"), overwrite=TRUE)  
```

#### Topographic Roughness  

```{r}
rough <- tri(elev, exact=FALSE)
names(rough) <- "tri"
writeRaster(rough, file.path(dat.dir.ex, "tri.tif"), overwrite=TRUE) 
```


#### Surface Relief Ratio 3x3 matrix

```{r}
rr <- srr(elev, 3) 
names(rr) <- "srr"
writeRaster(rr, file.path(dat.dir.ex, "srr3.tif"), overwrite=TRUE)
```


#### topographic position

```{r}
tp <- tpi(elev, 3)
names(tp) <- "tpi"
writeRaster(tp, file.path(dat.dir.ex, "tpi3.tif"),overwrite=TRUE)  
```

# Slope * COS(Aspect)

```{r}
sa <- terra::terrain(elev, v=c("slope", "aspect"), unit="degrees")
  scosa <- terra::lapp(c(sa[[1]], sa[[2]]), fun = sa.trans)
    names(scosa) <- "scosa"
      writeRaster(scosa, file.path(dat.dir.ex, "scosa.tif"), 
	              overwrite=TRUE)  
```

```{r}
#library(sf)
#wgs84 <- st_crs("EPSG:4326")
#htrs96 <- st_crs("EPSG:3765")
#etrs <- st_crs("EPSG:3035")
#dsf <- st_transform(st_as_sf(d, coords = c("xcoord", "ycoord"),crs = htrs96), etrs)    
```

